A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-03-03 podc_paper/introduction.tex
Adding some recent references

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations  in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation in a system of $p$ processes.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations  in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

### 2023-03-03 podc_paper/related.tex
Adding some recent references

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology
on array-based queues.
In worst-case executions that use the slow path,
they also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

@@

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free polylogarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.

>>>>

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
(It was shown recently that this implementation can be modified to use single-word CAS instructions rather than double-width CAS \cite{RK23}.)
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology
on array-based queues.
In worst-case executions that use the slow path,
they also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

@@

\emph{Restricted Queues.}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free polylogarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
Concurrently with our work, Johnen, Khattabi and Milani \cite{JKM23} build on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue implementation 
that achieves $O(\log p)$ time for enqueue operations and $O(k \log p)$ time for dequeue operations if there are only $k$ dequeuers.
